Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2018-03-05 paper/paper.tex
copyright
hogg stuff

% This project is part of the photoParallax project.
% Copyright 2017 the authors.

@@

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be precise and purely astrometric.
\gaia\ will use astrometric parallaxes to determine the distances of the more
precise stars, and then calibrate spectrophotometric models (\citealt{apsis}).
These spectrophotometric models, along with \gaia's on-board
low-resolution $BP/RP$ spectrophotometry, can be used to provide
more precise parallax estimates for stars with poor astrometric parallax data.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust, in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

Here we build and use just such a data-driven model to infer more precise parallaxes for each star.
In particular, we use all of the \gaia\ \tgas\ data (\citealt{tgas};
these data were generated by the methodology of \citealt{michalik15}), that match to the
\tmass\ photometry (\citealt{skrutskie06}) and lie in the footprint of \panstarrs\ (\psone; \citealt{ps1})
to make use of the \cite{green15} dust map. We use this data to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We build the \cmd\ using an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its disadvantage, for the present purposes, is that it requires
or implies a strictly Gaussian noise model.
As we show below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

Although not a fully Bayesian hierarchical inference, the methodology presented in this paper makes exceedingly weak assumptions
about the Milky Way, and similarly weak assumptions about the properties of
stars.
To date, there has been some discussion in the literature
about how to use a parallax responsibly to infer a distance (\citealt{astraatmadja16a}; \citealt{astraatmadja16b}).
These papers use a \gaia\ measurement to construct a parallax likelihood,
and combine this with a parallax prior.
While this is sensible and correct, all present methods proposed along
these lines build in informative (and known-to-be-wrong) assumptions
about the line-of-sight parallax distribution to \gaia\ stars.
That is, they build in strong and informative assumptions about the Milky
Way.
The strongest new assumption made in this work is that for every star
in \tgas, there are other, photometrically (and bolometrically)
similar stars.
That is a much weaker assumption than has been made in most other Bayesian
distance estimations.

>>>>

% This project is part of the photoParallax project.
% Copyright 2017, 2018 the authors.

@@

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be precise and purely astrometric
\added{The long-term complete distance information coming from \gaia\ will
involve astrometric, photometric, and spectral information, and the balance
of these contributions will be a strong function of stellar brightness (\citealt{apsis}).}
These spectrophotometric models, along with \gaia's on-board
low-resolution $BP/RP$ spectrophotometry, can be used to provide
more precise parallax estimates for stars with poor astrometric parallax data.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust, in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

Here we build and use just such a data-driven model to infer more precise parallaxes for each star.
In particular, we use all of the \gaia\ \tgas\ data (\citealt{tgas};
these data were generated by the methodology of \citealt{michalik15}), that match to the
\tmass\ photometry (\citealt{skrutskie06}) and lie in the footprint of \panstarrs\ (\psone; \citealt{ps1})
to make use of the \cite{green15} dust map. We use this data to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We build the \cmd\ using an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
\added{It also handles incomplete data, in the sense of there being missing
data values (these are like data values with arbitrarily large uncertainties);
this aspect of XD is not so relevant to this project,
where we have an essentially complete catalog.
(There is another sense for the word ``incomplete'' in this area, which has to do with
the completeness or selection function of the survey as a whole. We address this
kind of completeness in our assumption called ``selection'' above and in the Discussion.)}
XD's disadvantage, for the present purposes, is that it requires
or implies a strictly Gaussian noise model.
As we show below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

@@

Although not a fully Bayesian hierarchical inference, the methodology presented in this paper makes exceedingly weak assumptions
about the Milky Way, and similarly weak assumptions about the properties of
stars.
To date, there has been some discussion in the literature
about how to use a parallax responsibly to infer a distance (\citealt{astraatmadja16a}; \citealt{astraatmadja16b}).
These papers use a \gaia\ measurement to construct a parallax likelihood,
and combine this with a parallax prior.
While this is sensible and correct, all present methods proposed along
these lines build in informative (and known-to-be-wrong) assumptions
about the line-of-sight parallax distribution to \gaia\ stars.
That is, they build in strong and informative assumptions about the Milky
Way.
The \added{fundamental and critical} assumption made in this work is that for every star
in \tgas, there are other, photometrically (and bolometrically)
similar stars.
That is a much weaker assumption than has been made in most other Bayesian
distance estimations.
